\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2023-10-23 paper.tex
Ana's updates

In many settings, nodes are dated, i.e.\ each
node $u\in N$ is associated with a time $\tau_u$.
The topological relationships
defined by the directed graph structure, however,
 is sufficient for many applications,
and we will assume that node dates are not known here.
% This should be obvious, but seems like it needs to be said
In practical settings, we will wish to associate additional
metadata such as sample identifiers or quality-control metrics
with nodes; it is therefore best to think of the
integers used here in the definition of a node as an \emph{identifier},
with which arbitrary additional information can be associated.Concretely defining nodes as integers in this way has some
significant advantages. In particular, it provides a direct link
between an ARG and its local trees in terms of a well-defined
combinatorial object (Section~\ref{sec-ARG-and-local-trees}),
as well as enabling straightforward connection between
definitions and practical implementation (Section~\ref{sec-efficiency}).

@@

As illustrated in Fig.~\ref{fig-arg-in-pedigree},
the gARG for a given set of individuals is embedded in their pedigree.
The pedigree relationships between the diploid individuals is
shown, along with their paths
of genetic inheritance and constituent genomes (labelled
with lowercase alphabetical letters
rather than integer identifiers to avoid confusion with genomic
intervals).
Thus, individual $D_1$ is composed
of genomes \noderef{a} and \noderef{b}, which are inherited from its
two parents, $D_3$ and $D_4$. Each inherited genome may be the recombined product
of the genomes
% FIXME "contained within" is a poor phrase here, use "constituent" or something
contained within a parent.
In this example,
genome \noderef{b} was inherited directly from $D_4$'s genome \noderef{g} without
recombination, whereas
genome \noderef{a} is the recombinant product of
$D_2$'s genomes \noderef{e} and \noderef{f} crossing over at position 2.
Specifically, genome \noderef{a} inherited the (half-closed)
interval $[0, 2)$ from genome \noderef{e} and $[2, 10)$ from genome \noderef{f}.
These intervals are shown attached to the corresponding graph edges.
The figure shows the annotated pedigree with realised inheritance of genomes
between generations (A), the corresponding gARG (B), and finally the corresponding
sequence of local trees  along the
genome (C).
% Jerome: this feels unecessary and "genealogy" is often interpreted as
% "pedigree" so needs to be used very carefully.
% The local trees represent the genealogy of genomes \noderef{a}--\noderef{d},
% which changes between local trees due to recombination.
The local trees span the three genome regions delineated
by the two recombination breakpoints that gave rise to these genomes.
See Section~\ref{sec-ARG-and-local-trees} for further details on
how the local trees are embedded in an ARG.
% Following the above definitions, corresponding data encoding for
% this gARG is shown in Tab.~\ref{tab-gARG-data}.

@@

While the Griffiths approach of annotating recombination nodes
with a breakpoint in an eARG is a concise and elegant way of describing
realisations of the coalescent, it has some significant
issues when applied to the complexities of present-day datasets.
Because we are explicitly modelling the \emph{properties} of events,
the encoding must be extended to incorporate
different types of event that are not simple crossover recombination
or common ancestry.
Processes such as gene conversion, for example, can in principle
be included, but either require additional node types
or (e.g.) multiple recombination nodes joined by a zero-length edge.
Modelling multiple recombinant offspring of the same parents
(while keeping track of their identity as siblings)
as occurs
in Wright-Fisher models (e.g.\ Fig.~\ref{fig-arg-in-pedigree}),
or densely sampled multi-generational
data~\citep[e.g.][]{turnbull2018hundred,hayes20191000,RosFreixedes2022}
% circumlocutions isn't technically correct, but I like the roundabout
% overly verbose feel it gives.
would require more involved circumlocutions.
These and other problems can be worked around
(although, we would ideally like everyone to use the
\emph{same} workarounds if methods are to interoperate; see the Discussion),
but it is much simpler to use the gARG encoding described
in the previous section.
By recording the \emph{outcome} of events in terms of the
intervals of genetic inheritance on edges instead of recording the
properties of the events themselves, we remove the need to explicitly
model different types of event in the data structure.
More fundamentally, because gARGs
allow us to directly describe
genetic inheritance in multiple non-contiguous genome intervals
between an ancestor and a descendant,
we can encode the outcome of
\emph{multiple} genetic inheritance events, and to
systematically describe the uncertainty around their
ordering (see Section~\ref{sec-precision}).

>>>>

In many settings, nodes are dated, i.e.\ each
node $u\in N$ is associated with a time $\tau_u$.
The topological relationships
defined by the directed graph structure, however,
 is sufficient for many applications,
and we will assume that node dates are not known here.
% This should be obvious, but seems like it needs to be said
In practical settings, we will wish to associate additional
metadata such as sample identifiers or quality-control metrics
with nodes; it is therefore best to think of the
integers used here in the definition of a node as an \emph{identifier},
with which arbitrary additional information can be associated. Concretely defining nodes as integers in this way has some
significant advantages. In particular, it provides a direct link
between an ARG and its local trees in terms of a well-defined
combinatorial object (Section~\ref{sec-ARG-and-local-trees}),
as well as enabling straightforward connection between
definitions and practical implementation (Section~\ref{sec-efficiency}).

@@

As illustrated in Fig.~\ref{fig-arg-in-pedigree},
the gARG for a given set of individuals is embedded in their pedigree.
The pedigree relationships between the diploid individuals are
shown, along with their paths
of genetic inheritance and constituent genomes (labelled
with lowercase alphabetical letters
rather than integer identifiers to avoid confusion with genomic
intervals).
Thus, individual $D_1$ is composed
of genomes \noderef{a} and \noderef{b}, which are inherited from its
two parents, $D_3$ and $D_4$. Each inherited genome may be the recombined product
of the genomes
% FIXME "contained within" is a poor phrase here, use "constituent" or something
contained within a parent.
In this example,
genome \noderef{b} was inherited directly from $D_4$'s genome \noderef{g} without
recombination, whereas
genome \noderef{a} is the recombinant product of
$D_2$'s genomes \noderef{e} and \noderef{f} crossing over at position 2.
Specifically, genome \noderef{a} inherited the (half-closed)
interval $[0, 2)$ from genome \noderef{e} and $[2, 10)$ from genome \noderef{f}.
These intervals are shown attached to the corresponding graph edges.
The figure shows the annotated pedigree with realised inheritance of genomes
between generations (A), the corresponding gARG (B), and finally the corresponding
sequence of local trees  along the
genome (C).
% Jerome: this feels unecessary and "genealogy" is often interpreted as
% "pedigree" so needs to be used very carefully.
% The local trees represent the genealogy of genomes \noderef{a}--\noderef{d},
% which changes between local trees due to recombination.
The local trees span the three genome regions delineated
by the two recombination breakpoints that gave rise to these genomes.
See Section~\ref{sec-ARG-and-local-trees} for further details on
how the local trees are embedded in an ARG.
% Following the above definitions, corresponding data encoding for
% this gARG is shown in Tab.~\ref{tab-gARG-data}.

@@

While the Griffiths approach of annotating recombination nodes
with a breakpoint in an eARG is a concise and elegant way of describing
realisations of the coalescent, it has some significant
issues when applied to the complexities of present-day datasets.
Because we are explicitly modelling the \emph{properties} of events,
the encoding must be extended to incorporate
different types of event that are not simple crossover recombination
or common ancestry.
Processes such as gene conversion, for example, can in principle
be included, but either require additional node types
or, for instance, multiple recombination nodes joined by a zero-length edge.
Modelling multiple recombinant offspring of the same parents
(while keeping track of their identity as siblings)
as occurs
in Wright-Fisher models (e.g.\ Fig.~\ref{fig-arg-in-pedigree}),
or densely sampled multi-generational
data~\citep[e.g.][]{turnbull2018hundred,hayes20191000,RosFreixedes2022}
% circumlocutions isn't technically correct, but I like the roundabout
% overly verbose feel it gives.
would require more involved circumlocutions.
These and other problems can be worked around
(although, we would ideally like everyone to use the
\emph{same} workarounds if methods are to interoperate; see the Discussion),
but it is much simpler to use the gARG encoding described
in the previous section.
By recording the \emph{outcome} of events in terms of the
intervals of genetic inheritance on edges instead of recording the
properties of the events themselves, we remove the need to explicitly
model different types of event in the data structure.
More fundamentally, because gARGs
allow us to directly describe
genetic inheritance in multiple non-contiguous genome intervals
between an ancestor and a descendant,
we can encode the outcome of
\emph{multiple} genetic inheritance events, and to
systematically describe the uncertainty around their
ordering (see Section~\ref{sec-precision}).
